---
title: "Untitled"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages

```{r}
library(readxl)
library(readr)
library(openxlsx)
library(latex2exp)
library(xtable)
library(agricolae)
library(forcats)
library(tidyverse)
```

## Simulation examples and some tests

simulations test.

```{r}
set.seed(12345)

m1 <- m2 <- 2
m3 <- 10
n1 <- n2 <- 100
n3 <- 200
dados <- tibble(y = c(rnorm(n1, m1), rnorm(n2, m2), rnorm(n3, m3)),
                x = c(rep("m1", n1), rep("m2", n2), rep("m3", n3)))
dados <- dados %>% 
  mutate(x = as_factor(x))

ajuste <- aov(y ~ x, data = dados)

tt <- LSD.test(ajuste, "x", console = T)

tt
  
ggplot(dados) + 
  geom_boxplot(aes(y = y, x = x))
```

## Graphs illustrating the t-test and z-test

Duas populações.

```{r}
z <- seq(from = -3, to = 3, by = 0.0001)
m1 <- 5
m2 <- 15
x1 <- z + m1
y1 <- dnorm(x1, mean = m1)
x2 <- z + m2
y2 <- dnorm(x2, mean = m2)
dados <- tibble(x1, x2, media1 = y1, media2 = y2)

ggplot(dados) +
  geom_line(aes(x = x1, y = media1), size = 2, color = "blue") +
  geom_line(aes(x = x2, y = media2), size = 2, color = "blue") +
  geom_vline(xintercept = m1, size = 1) +
  geom_vline(xintercept = m2, size = 1) +
  geom_segment(x = min(c(x1, x2)), y = 0, xend = max(c(x1, x2)), yend = 0) +
  labs(x = "", y = "") +
  annotate("text", x = min(x1) * 1.5, y = dnorm(m1, mean = m1), label = "Tratamento 1", size = 5)+
  annotate("text", x = min(x2) * 1.05, y = dnorm(m2, mean = m2), label = "Tratamento 2", size = 5)+
  theme_minimal() +
  scale_x_continuous(breaks = c(m1, m2),
                     labels = c(expression(mu[1]), expression(mu[2])))+
  scale_y_continuous(breaks = NULL) +
  labs(title = bquote("Distribuição normal"),
       subtitle = bquote("Variância: " ~ sigma^2)) +
  theme(axis.text = element_text(size = 25), 
        title = element_text(size = 25))
ggsave("../figure/pop-2.png")
ggsave("../figure/pop-2.pdf")


```

Let's illustrate the boxplots with a simulated example.
```{r}
set.seed(1235813)
m1 <- 5
m2 <- 15
n <- 1e+3

dados <- tibble(
  x = c(rep("Tratamento 1", n), rep("Tratamento 2", n)),
  y = c(rnorm(n, mean = m1), rnorm(n, mean = m2))
)
ggplot(dados) +
  geom_boxplot(aes(x = x, y = y), outlier.shape = NA) +
  labs(x = "", y = "", title = "Diagrama de caixa") +
  scale_y_continuous(breaks = c(m1, m2),
                     labels = c(
                       expression(mu[1]),
                       expression(mu[2])
                     ))+
  theme_classic() +
  theme(axis.text = element_text(size = 25),
        title = element_text(size = 30))
ggsave("../figure/boxplot-pop-2.png")
ggsave("../figure/boxplot-pop-2.pdf")


```


Três populações.

```{r}
z <- seq(from = -3, to = 3, by = 0.0001)
m1 <- 5
m2 <- 15
m3 <- 25
x1 <- z + m1
y1 <- dnorm(x1, mean = m1)
x2 <- z + m2
y2 <- dnorm(x2, mean = m2)
x3 <- z + m3
y3 <- dnorm(x3, mean = m3)
dados <- tibble(x1, x2, x3, media1 = y1, media2 = y2, media3 = y3)

ggplot(dados) +
  geom_line(aes(x = x1, y = media1), size = 2, color = "blue") +
  geom_line(aes(x = x2, y = media2), size = 2, color = "blue") +
  geom_line(aes(x = x3, y = media3), size = 2, color = "blue") +
  geom_vline(xintercept = m1, size = 1) +
  geom_vline(xintercept = m2, size = 1) +
  geom_vline(xintercept = m3, size = 1) +
  geom_segment(x = min(c(x1, x2, x3)), y = 0, xend = max(c(x1, x2, x3)), yend = 0) +
  labs(x = "", y = "") +
  annotate("text", x = min(x1) * 1.3, y = dnorm(m1, mean = m1), label = "Tratamento 1", size = 4.5)+
  annotate("text", x = min(x2) * 1.0, y = dnorm(m2, mean = m2), label = "Tratamento 2", size = 4.5)+
  annotate("text", x = min(x3) * 1.0, y = dnorm(m3, mean = m3), label = "Tratamento 3", size = 4.5)+
  theme_minimal() +
  scale_x_continuous(breaks = c(m1, m2, m3),
                     labels = c(expression(mu[1]), 
                                expression(mu[2]),
                                expression(mu[3])))+
  scale_y_continuous(breaks = NULL) +
  labs(title = bquote("Distribuição normal"),
       subtitle = bquote("Variância: " ~ sigma^2)) +
  theme(axis.text = element_text(size = 25), 
        title = element_text(size = 25))
ggsave("../figure/pop-3.png")
ggsave("../figure/pop-3.pdf")

```

Let's illustrate the boxplots with a simulated example.
```{r}
set.seed(1235813)
m1 <- 5
m2 <- 15
m3 <- 25
n <- 1e+3

dados <- tibble(
  x = c(rep("Tratamento 1", n), rep("Tratamento 2", n), rep("Tratamento 3", n)),
  y = c(rnorm(n, mean = m1), rnorm(n, mean = m2), rnorm(n, mean = m3))
)
ggplot(dados) +
  geom_boxplot(aes(x = x, y = y),outlier.shape = NA) +
  labs(x = "", y = "", title = "Diagrama de caixa") +
  scale_y_continuous(breaks = c(m1, m2, m3),
                     labels = c(
                       expression(mu[1]),
                       expression(mu[2]),
                       expression(mu[3])
                     ))+
  theme_classic() +
  theme(axis.text = element_text(size = 25),
        title = element_text(size = 30))
ggsave("../figure/boxplot-pop-3.png")
ggsave("../figure/boxplot-pop-3.pdf")


```

## Graph with decision idea

```{r}
df1 = 4
df2 = 20
xmax <- 8
z <- seq(from = 0, to = xmax, by = 0.0001)
y <- df(z, df1 = df1, df2 = df2)
dados <- tibble(z = z, y = y)
a <- 0.05
q <- qf(1 - a, df1 = df1, df2 = df2)

ggplot(dados) +
  geom_line(aes(x = z, y = y), color = "black", size = 1) +
  labs(x = TeX("$F_0$"), y = "Função densidade", 
       title = TeX("Rejeitamos $H_0: \\tau_{1} = \\tau_{2} = \\cdots = \\tau_{a} = 0$ se $f_0 > f_{1-\\alpha; a-1, N-a}$"),
       subtitle = bquote("RC = {" ~ f[0] ~ " | " ~ f[0] > f[1-alpha~";"~a-1~","~N-a] ~ "}"),
       caption = bquote(f[0] == frac(QM[Tratamentos], QM[E]) ~"e"~N == n %.% a)) +
  scale_x_continuous(breaks = c(q),
                     labels = c(TeX("$f_{1-\\alpha; a-1, N-a}$"))) +
  scale_y_continuous(breaks = NULL) +
  geom_ribbon(data = dplyr::filter(dados, z > q),
              aes(x = z, y = y, ymin = 0, ymax = y),
              fill = "gray") +
  theme_classic() +
  theme(axis.title = element_text(size = 30),
        axis.text = element_text(size = 25),
        title = element_text(size = 20)) +
  geom_segment(x = -0.5, xend = q, y = 0.15, yend =0.15, size = 1, color = "blue") +
  geom_segment(x = q, xend = xmax, y = 0.15, yend =0.15, size = 1, color = "red") +
  geom_segment(x = q, xend = q, y = 0, yend = 0.15,
               linetype = "dashed", color = "gray", size = 1) +
  geom_point(data = tibble(x = c(q), y = c(0.15)),
             aes(x = x, y = y), size = 5,color  = "gray") +
  annotate("text", x = 1, y = 0.2, label = "H[0]",
           size = 10, color = "blue", parse = T) +
  annotate("text", x = 3.5, y = 0.2, label = "H[1]",
           size = 10, color = "red", parse = T) +
  annotate("text", x = 3.5, y = 0.075, label = "alpha", parse = T,
           color = "gray", size = 10)
ggsave("../figure/anova-test.png")
ggsave("../figure/anova-test.pdf")

```


## Example

```{r}
set.seed(1235813)

dados <- tribble(~`Algodão`, ~o1, ~o2, ~o3, ~o4, ~o5, ~Total, ~`Variância`,
                  "15",  7L, 7L, 15L, 11L, 9L, NA , NA,
                  "20", 12L, 17L,  12L, 18L, 18L, NA , NA,
                 "25",  14L, 18L, 18L, 19L, 19L, NA, NA,
                 "30",  19L, 25L, 22L, 19L, 23L, NA, NA,
                 "35",  7L, 10L, 11L, 15L, 11L, NA, NA
)

dados <- dados %>% group_by(`Algodão`) %>% 
  mutate(Total = sum(c(o1, o2, o3, o4, o5)),
         `Variância` = var(c(o1, o2, o3, o4, o5)),
         `Média` = mean(c(o1, o2, o3, o4, o5)))

dados %>% 
  xtable(caption = "Resistência à tração", 
         label = "tab:resistencia",
         digits = 2,
         align = rep("c", ncol(dados) + 1)) %>% 
  print.xtable(booktabs = T,
               include.rownames = F,
               format.args = list(decimal.mark = ","))

dados1 <- dados %>% select(-Total,-`Variância`) %>% 
  tidyr::gather(key = "ordem", value = "obs", str_c("o", 1:5)) %>% 
  select(-ordem) %>% ungroup() %>% 
  mutate(`Algodão` = forcats::as_factor(`Algodão`))

dados1 %>% 
  summarise(media = mean(obs), variancia = var(obs)) %>% 
  xtable(digits = 2,
         caption = "Média e variância de todas observações de todos os tratamentos.",
         label = "tab:media-variancia-pop") %>% 
  print.xtable(booktabs = T,
               format.args = list(decimal.mark = ","),
               include.rownames = F)

dados1 %>% group_by(`Algodão`) %>% 
  summarise(`Médias populacionais` = mean(obs) %>% round()) %>% 
  xtable(caption = "Médias populacionais.",
         digits = 0,
         label = "tab:medias-tratamentos") %>% 
  print.xtable(booktabs = T,
               format.args = list(decimal.mark = ","),
               include.rownames = F)


aov(obs ~ `Algodão`, data = dados1) %>%
  summary() %>%
  xtable(digits = 2, caption = "Tabela ANOVA.", label="tab:anova-table-ex1") %>%
  print.xtable(booktabs = T, format.args = list(decimal.mark = ","))

# análise de resíduos 
ajuste <- aov(obs ~ `Algodão`, data = dados1)
analise_residuos <- broom::augment(ajuste) 
analise_residuos <- analise_residuos %>% 
  select(obs, `Algodão`, .fitted, .resid)

analise_residuos %>% 
  filter(`Algodão` %in% 35) %>% 
  select(.resid) %>% 
  round(4) %>% 
  unlist() %>% 
  paste(collapse = " & ") %>% 
  str_replace_all("\\.", ",")

m <- analise_residuos %>% 
  group_by(Algodão) %>% 
  summarise(media = mean(obs)) %>% select(media) %>% ungroup() %>% 
  unlist()

with(analise_residuos, qqnorm((.resid - mean(.resid)) / sd(.resid), plot.it = F)) %>% 
  as_tibble() %>% 
  ggplot()+
  geom_point(aes(x, y), color = "blue", size = 3)+
  labs(x = "Quantis teóricos", y = "Quantis amostrais") +
  xlim(c(-3, 3)) +
  ylim(c(-3, 3)) +
  geom_abline(slope = 1, intercept = 0, size = 2, alpha = 0.75)+
  theme_minimal() +
  theme(axis.title = element_text(size = 30),
        axis.text = element_text(size = 25))
ggsave("../figure/qqnorm-balanced-ex1.pdf")
ggsave("../figure/qqnorm-balanced-ex1.png")

analise_residuos %>% 
  ggplot() +
  geom_point(aes(x = seq_along(.resid), y = .resid), size = 3, color = "blue") +
  labs(x = "Tempo", y = "Resíduos") +
  theme_minimal()+
  theme(axis.title = element_text(size = 30),
        axis.text = element_text(size = 25))
ggsave("../figure/independence-balanced-ex1.pdf")
ggsave("../figure/independence-balanced-ex1.png")

analise_residuos %>% 
  ggplot() +
  geom_point(aes(x = .fitted, y = .resid), size = 3, color = "blue") +
  labs(x = "Valor ajustado", y = "Resíduos") +
  scale_x_continuous(breaks = unique(analise_residuos$.fitted) %>% round(2)) +
  theme_minimal()+
  theme(axis.title = element_text(size = 30),
        axis.text.x = element_text(size = 15, angle = 60),
        axis.text.y = element_text(size = 15))
ggsave("../figure/variance-balanced-ex1.pdf")
ggsave("../figure/variance-balanced-ex1.png")

# boxplot de cada tratamento
dados1 %>% 
  group_by(`Algodão`) %>% 
  summarise(lower = quantile(obs, probs = 0.25),
            middle = quantile(obs, probs = 0.5),
            upper = quantile(obs, probs = 0.75),
            min = min(obs), max = max(obs), 
            li = lower - 1.5 * (upper - lower),
            ls = upper + 1.5 * (upper - lower)) %>%
  ggplot() +
  geom_boxplot(aes(x = `Algodão`,
                   lower = lower, 
                   middle = middle, 
                   upper = upper,
                   ymin = li, 
                   ymax = ls), stat = "identity")+
  labs(x = "Porcentagem de fibra de algodão",
       y = "Resistência à tração do tecido") +
  theme_classic()+
  theme(axis.title = element_text(size = 30),
        axis.text = element_text(size = 25))

ggsave("../figure/boxplot-ex1.pdf")
ggsave("../figure/boxplot-ex1.png")
```

Below, we compute the power and sample size.
```{r}
samp_size_anova <- function(a, es, alpha = 0.05, power = 0.95){
  
  f <- function(n){
    n %>%
      map_dbl(~ (
      1 - pf(qf(1 - alpha, df1 = a - 1, df2 = .x * a - a),
             df1 = a  - 1,
             df2 = .x * a - a,
             ncp = .x * es^2)-
        power)^2)
  }
  
  nlminb(2, f)$par %>% ceiling()
}
a <- 5
tau <- c(0, 5, 10, 15, 20)
s2 = 27
es <- (sum(tau^2) / (a * s2)) %>% sqrt()
samp_size_anova(a = 5, es = es, alpha = 0.05, power = 0.99)

```


## Graph with decision idea (unbalaced)

```{r}
df1 = 4
df2 = 20
xmax <- 8
z <- seq(from = 0, to = xmax, by = 0.0001)
y <- df(z, df1 = df1, df2 = df2)
dados <- tibble(z = z, y = y)
a <- 0.05
q <- qf(1 - a, df1 = df1, df2 = df2)

ggplot(dados) +
  geom_line(aes(x = z, y = y), color = "black", size = 1) +
  labs(x = TeX("$F_0$"), y = "Função densidade", 
       title = TeX("Rejeitamos $H_0: \\tau_{1} = \\tau_{2} = \\cdots = \\tau_{a} = 0$ se $f_0 > f_{1-\\alpha; a-1, N-a}$"),
       subtitle = bquote("RC = {" ~ f[0] ~ " | " ~ f[0] > f[1-alpha~";"~a-1~","~N-a] ~ "}"),
       caption = bquote(f[0] == frac(QM[Tratamentos], QM[E])~"e"~ 
                           N == n[1]+ cdots + n[a])) +
  scale_x_continuous(breaks = c(q),
                     labels = c(TeX("$f_{1-\\alpha; a-1, N-a}$"))) +
  scale_y_continuous(breaks = NULL) +
  geom_ribbon(data = dplyr::filter(dados, z > q),
              aes(x = z, y = y, ymin = 0, ymax = y),
              fill = "gray") +
  theme_classic() +
  theme(axis.title = element_text(size = 30),
        axis.text = element_text(size = 25),
        title = element_text(size = 20)) +
  geom_segment(x = -0.5, xend = q, y = 0.15, yend =0.15, size = 1, color = "blue") +
  geom_segment(x = q, xend = xmax, y = 0.15, yend =0.15, size = 1, color = "red") +
  geom_segment(x = q, xend = q, y = 0, yend = 0.15,
               linetype = "dashed", color = "gray", size = 1) +
  geom_point(data = tibble(x = c(q), y = c(0.15)),
             aes(x = x, y = y), size = 5,color  = "gray") +
  annotate("text", x = 1, y = 0.2, label = "H[0]",
           size = 10, color = "blue", parse = T) +
  annotate("text", x = 3.5, y = 0.2, label = "H[1]",
           size = 10, color = "red", parse = T) +
  annotate("text", x = 3.5, y = 0.075, label = "alpha", parse = T,
           color = "gray", size = 10)
ggsave("../figure/anova-test-unbalanced.png")
ggsave("../figure/anova-test-unbalanced.pdf")

```

## Unbalanced experiments

Below we present an example.
```{r}
temps <- c(100,
 125,
 150,
 175
 )
weathermetrics::fahrenheit.to.celsius(temps) %>% ceiling()

dados <- tribble(
  ~Densidade, ~Temperatura,
 21.8, 40,
 21.9, 40,
 21.7, 40,
 21.6, 40,
 21.7, 40,
 21.5, 40,
 21.8, 40,
 21.7, 50,
 21.4, 50,
 21.5, 50, 
 21.5, 50,
 21.9, 60,
 21.8, 60,
 21.8, 60,
 21.6, 60,
 21.5, 60, 
  21.9, 70,
 21.7, 70,
 21.8, 70,
 21.7, 70, 
 21.6, 70,
 21.8, 70
) %>% 
  transmute(Densidade, `Temperatura C` = forcats::as_factor(Temperatura))

aov(data = dados, Densidade ~ `Temperatura C`) %>% 
  xtable(caption = "Tabela Anova",
         digits = 3,
         label = "tab:anova-unbalanced-test") %>% 
  print.xtable(booktabs = T, format.args = list(decimal.mark = ","))

dados %>% group_by(`Temperatura C`) %>% 
  summarise(media = mean(Densidade), variancia = var(Densidade)) %>% 
  xtable(digits = 4) %>% 
  print.xtable(include.rownames = F, booktabs = T, format.args = list(decimal.mark = ","))

ajuste <- aov(data = dados, Densidade ~ `Temperatura C`)
analise_residuos <- broom::augment(ajuste) %>%
  select(Densidade, Temperatura.C, .fitted, .resid)

analise_residuos %>% 
  filter(Temperatura.C %in% 70) %>%  
  select(.resid) %>% 
  unlist() %>% 
  round(4) %>% 
  paste(collapse = " & ") %>% 
  str_replace_all("\\.", ",")

# qqnorm plot (ggplot2 style)
with(analise_residuos, qqnorm((.resid - mean(.resid)) / sd(.resid), plot.it = F)) %>% 
  as_tibble() %>% 
  ggplot()+
  geom_point(aes(x, y), color = "blue", size = 3)+
  labs(x = "Quantis teóricos", y = "Quantis amostrais") +
  xlim(c(-3, 3)) +
  ylim(c(-3, 3)) +
  geom_abline(slope = 1, intercept = 0, size = 2, alpha = 0.75)+
  theme_minimal() +
  theme(axis.title = element_text(size = 30),
        axis.text = element_text(size = 25))
ggsave("../figure/qqnorm-unbalanced.pdf")
ggsave("../figure/qqnorm-unbalanced.png")

analise_residuos %>% 
  ggplot() +
  geom_point(aes(x = seq_along(.resid), y = .resid), size = 3, color = "blue") +
  labs(x = "Tempo", y = "Resíduos") +
  theme_minimal()+
  theme(axis.title = element_text(size = 30),
        axis.text = element_text(size = 25))
ggsave("../figure/independence-unbalanced.pdf")
ggsave("../figure/independence-unbalanced.png")

analise_residuos %>% 
  ggplot() +
  geom_point(aes(x = .fitted, y = .resid), size = 3, color = "blue") +
  labs(x = "Valor ajustado", y = "Resíduos") +
  scale_x_continuous(breaks = unique(analise_residuos$.fitted) %>% round(2)) +
  theme_minimal()+
  theme(axis.title = element_text(size = 30),
        axis.text.x = element_text(size = 15, angle = 60),
        axis.text.y = element_text(size = 15))
ggsave("../figure/variance-unbalanced.pdf")
ggsave("../figure/variance-unbalanced.png")


```

